RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R

### 2015-04-27 manuscripts/manuscript.Rmd
Edits to the manuscript revision.

---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
  - name: Hilmar Lapp
    affiliation: NESCent
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and extensible exchange standard
      recently proposed to better meet the expanding needs for
      phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which provides users of
      the R programming language with easy-to-use tools for 
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

@@

A detailed account of how the NeXML standard addresses these and
other relevant challenges can be found in @Vos_2012. In brief,
NeXML was designed with the following important properties. First,
NeXML is defined by a precise grammar that can be programmatically
**validated**: meaning that we can verify that a file has followed this
precise grammar and thus be sure that it can be read (parsed) 
without errors by any software that uses the NeXML grammar (e.g. RNeXML). 
Second, NeXML is **extensible**: a user
can define representations of new, previously unanticipated information
(as we will illustrate) without violating its defining grammar. Third
and most importantly, NeXML is rich in **computable semantics**: patterns
that permit a machine to reason about concepts. Note that
these semantics are based on logical deduction such as: "all frogs are vertebrates,
so a query for vertebrates should return all species that are listed as frogs"), 
rather than heuristic machine learning algorithms.  Such inference requires a
computer to understand the 'semantic' meaning of a word like 'frog', which
in must be specified in a kind of dictionary known in informatics research
as an vocabulary or ontology (see @Parr2011 for a more detailed introduction
in the context of evolution and biodiversity research).

@@

The `RNeXML` interface described above for built-in metadata allows
users to create precise and semantically rich annotations without 
confronting any of the complexity of namespaces and ontologies.
Nevertheless, advanced users may desire the explicit control over
these semantic tools that takes full advantage of the flexibility
and extensibility of the NeXML specification [@Vos2012; @Parr2011].
In this section we detail how to accomplish these more complex uses
in RNeXML.

Using a vocabulary or ontology terms rather than simple text strings to 
describe data is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` interface shields her otherwise: in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

NEXML's extensibility avoids this problem.  In NeXML, it is perfectly
possible to add very complex annotations while maintaining a format that
is valid NeXML syntax which can be read by any NeXML parser. To illustrate
how RNeXML facilitates extending the NeXML standard to accommodate such 
complex annotations, we have implemented two functions in the package,
`nexml_to_simmap` and `simmap_to_nexml`.  These functions illustrate
how simmap data can be represented as `meta` annotations on the branch
length elements of a NeXML tree, and provide routines to convert between
this NeXML representation and the extended `ape::phylo` representation
of a `simmap` tree in R that was introduced by @Revell_2012. We encourage
readers to consult the example code in `simmap_to_nexml` to see how
this is implemented.

>>>>

---
#layout: preprint
layout: review, 11pt
linenumbers: true
title: "RNeXML: a package for reading and writing richly annotated phylogenetic, character, and trait data in R"
author: 
  - name: Carl Boettiger
    affiliation: cstar
    email: cboettig(at)gmail.com
    footnote: Corresponding author
  - name: Scott Chamberlain
    affiliation: ropensci
  - name: Rutger Vos
    affiliation: NBC
  - name: Hilmar Lapp
    affiliation: dukeplus
address: 
  - code: cstar
    address: | 
      Center for Stock Assessment Research, 
      Department of Applied Math and Statistics, 
      University of California, Mail Stop SOE-2,
      Santa Cruz, CA 95064, USA
  - code: NESCent
    address: | 
      Center for Genomic and Computational Biology, Duke University, Durham, NC, USA
      and 
      National Evolutionary Synthesis Center and Duke University, Durham, NC, USA
  - code: ropensci
    address: |
      University of California, Berkeley, CA, USA
  - code: NBC
    address: | 
      Naturalis Biodiversity Center, Leiden, the Netherlands
abstract: | 
      1. NeXML is a powerful and extensible exchange standard
      recently proposed to better meet the expanding needs for
      phylogenetic data and metadata sharing. Here we
      present the RNeXML package, which provides users of
      the R programming language with easy-to-use tools for 
      reading and writing NeXML documents, including rich metadata, in
      a way that interfaces seamlessly with the extensive library of
      phylogenetic tools already available in the R ecosystem.

@@

A detailed account of how the NeXML standard addresses these and
other relevant challenges can be found in @Vos_2012. In brief,
NeXML was designed with the following important properties. First,
NeXML is defined by a precise grammar that can be programmatically
**validated**; i.e., it can be verified whether a file precisely
follows this grammar, and therefore whether it can be read (parsed) 
without errors by software that uses the NeXML grammar (e.g. RNeXML)
is predictable. 
Second, NeXML is **extensible**: a user
can define representations of new, previously unanticipated information
(as we will illustrate) without violating its defining grammar. Third
and most importantly, NeXML is rich in **computable semantics**: it is
designed for expressing metadata such that machines can understand
their meaning and make inferences from it. For example, OTUs in a tree
or character matrix for frog species can be linked to concepts in a
formally defined hierarchy of taxonomic concepts such as the
Vertebrate Taxonomy Ontology [@Midford_2014], which enables a machine
to infer that a query for amphibia is to include the frog data in what
is returned. (For a more broader discussion of the value of such
capabilities for evolutionary and biodiversity science we refer the
reader to @Parr2011.)

@@

The `RNeXML` interface described above for built-in metadata allows
users to create precise and semantically rich annotations without 
confronting any of the complexity of namespaces and ontologies.
Nevertheless, advanced users may desire the explicit control over
these semantic tools to take full advantage of the flexibility
and extensibility of the NeXML specification [@Vos2012; @Parr2011].
In this section we detail how to accomplish these more complex uses
in RNeXML.

Using a vocabulary or ontology terms rather than simple text strings to 
describe data is crucial for allowing
machines to not only parse but also interpret and potentially reason
over their semantics. To achieve this benefit for custom metadata
extensions, the user necessarily needs to handle certain technical
details from which the `RNeXML` interface shields her otherwise, in particular
the globally unique identifiers (normally HTTP URIs) of metadata terms
and vocabularies. To be consistent with XML terminology, `RNeXML` calls
vocabulary URIs _namespaces_, and their abbreviations _prefixes_. For
example, the namespace for the Dublin Core Metadata Terms vocabulary
is "http://purl.org/dc/elements/1.1/". Using its common abbreviation
"dc", a metadata property "dc:title" expands to the identifier
"http://purl.org/dc/elements/1.1/title". This URI resolves to a human and
machine-readable (depending on access) definition of precisely what the
term `title` in Dublin Core means. In contrast, just using the text string
"title" could also mean the title of a person, a legal title, the verb
title, etc.  URI identifiers of metadata vocabularies and terms are not
mandated to resolve, but if machines are to derive the maximum benefit
from them, they should resolve to a definition of their semantics in RDF.

@@

In contrast, by allowing new data types to be added as --- sometimes
complex --- metadata annotations NeXML can accommodate data extensions
without compromise to its grammar and thus syntax In NeXML. To illustrate
how RNeXML facilitates extending the NeXML standard in this way, we
have implemented two functions in the package, `nexml_to_simmap` and
`simmap_to_nexml`.  These functions show
how simmap data can be represented as `meta` annotations on the branch
length elements of a NeXML tree, and provide routines to convert between
this NeXML representation and the extended `ape::phylo` representation
of a `simmap` tree in R that was introduced by @Revell_2012. We encourage
readers interested in this capability to consult the example code in
`simmap_to_nexml` to see how this is implemented.
